Making Sequential Programs Deterministic

### 2019-04-05 design.tex
design minor edits

It is possible to store an expression of type \<Det> in a \<NonDet>
collection, but any value retrieved from a \<NonDet> collection has
type \<NonDet>.

@@

\noindent
and no type introduced so far can express this polymorphism.
(The typing is actually even more complex, because if \emph{either}
argument to \<get> is \<OrderNonDet> or \<NonDet>, then the
result is \<NonDet>.  The above examples only show the case where the index
is deterministic.)
The typical type fro the list access operation only applies if both
arguments are \<Det>,
and our typing must have that as a special case, which is the first line above.

@@

\noindent
because such an instantiation would include the invalid return type \<OrderNonDet Int>.
This means that the only two instantiations are at 
$\kappa = \<NonDet>$ and $\kappa = \<Det>$.  An invocation on a
\<OrderNonDet> collection would choose the \<NonDet> instantiation and the
return type at that call site would be \<NonDet Int> rather than the
desired \<Det Int>.

@@

$\<listOfLists> : \ptype{OND List}{\ptype{Det List}{Det String}}$

@@

The core calculus requires a qualifier on every type.}

@@

\TheDeterminismChecker has additional type refinement rules for the sorting
methods of \<List>, \<Collections>, and \<Arrays> classes
\begin{figure}
%     $\infer[\rulename{list sort}]{\Gamma \vdash x : \|Det \ List|\angles{\<Det> \ \beta}}{\Gamma \vdash x : \|OrderNonDet \ List|\angles{\<Det> \ \beta} & \|x.sort()|}$
%     
%     \bigskip

@@

We annotated the \codeid{equals()} method of the \<Set> class in the JDK as
\begin{verbatim}
PolyDet("down") boolean equals(PolyDet Set<E> this, PolyDet Object o)
\end{verbatim}
While this is sound, it is imprecise in the case of \codeid{OrderNonDet
  Set}s.\todo{Explain why.}
 We added the following rule shown in \cref{fig-precision-rules} to our checker to improve precision.
\begin{figure}
    \todo{add line break}
    $\infer[\rulename{set precision}]{\Gamma \vdash \|x.equals(y)| : \<Det>}{\Gamma \vdash \|x| : \<OrderNonDet> \ \|Set|\angles{ \kappa_{x} \ \beta_{x}}, \|y| : \<OrderNonDet> \ \|Set|\angles{ \kappa_{y} \ \beta_{y}} & \kappa_{x} \ \beta_{x} \nsqsubseteq \<OrderNonDet> \ \<List> & \kappa_{y} \ \beta_{y} \nsqsubseteq \<OrderNonDet> \ \<List> }$

>>>>

%It is possible to store an expression of type \<Det> in a \<NonDet>
%collection, but any value retrieved from a \<NonDet> collection has
%type \<NonDet>.

@@

\noindent
and no type introduced so far can express this polymorphism.
(The typing is actually even more complex, because if \emph{either}
argument to \<get> is \<OrderNonDet> or \<NonDet>, then the
result is \<NonDet>.  The above examples only show the case where the index
is deterministic.)
The typical type for the list access operation only applies if both
arguments are \<Det>,
and our typing must have that as a special case, which is the first line above.

@@

\noindent
because such an instantiation would include the invalid return type \<OrderNonDet Int>.
This means that the only two instantiations are at 
$\kappa = \<NonDet>$ and $\kappa = \<Det>$.  An invocation on an
\<OrderNonDet> collection would choose the \<NonDet> instantiation and the
return type at that call site would be \<NonDet Int> rather than the
desired \<Det Int>.

@@

$\<listOfLists> : \ptype{OrderNonDet List}{\ptype{Det List}{Det String}}$

@@

\todo{The core calculus requires a qualifier on every type.}

@@

\TheDeterminismChecker has additional type refinement rules for the sorting
methods of \<List>, \<Collections>, and \<Arrays> classes.
\begin{figure}
%     $\infer[\rulename{list sort}]{\Gamma \vdash x : \|Det \ List|\angles{\<Det> \ \beta}}{\Gamma \vdash x : \|OrderNonDet \ List|\angles{\<Det> \ \beta} & \|x.sort()|}$
%     
%     \bigskip

@@

We annotated the \codeid{equals()} method of the \<Set> class in the JDK as
\begin{verbatim}
PolyDet("down") boolean equals(PolyDet Set<E> this, PolyDet Object o)
\end{verbatim}
While this is sound, it is imprecise in the case of \codeid{OrderNonDet
  Set}s.\todo{Explain why.}
 We added the rule shown in \cref{fig-precision-rules} to our checker to improve precision.
\begin{figure}
    \todo{add line break}
    $\infer[\rulename{set precision}]{\Gamma \vdash \|x.equals(y)| : \<Det>}{\Gamma \vdash \|x| : \<OrderNonDet> \ \|Set|\angles{ \kappa_{x} \ \beta_{x}}, \|y| : \<OrderNonDet> \ \|Set|\angles{ \kappa_{y} \ \beta_{y}} & \kappa_{x} \ \beta_{x} \nsqsubseteq \<OrderNonDet> \ \<List> & \kappa_{y} \ \beta_{y} \nsqsubseteq \<OrderNonDet> \ \<List> }$
